### 2013-06-10 doc/dissertation/document/chapters/mc_background/mc_background.tex
restructuring verification results
mc background edits and spell check

\chapter{Monte Carlo Solution Methods for Linear Systems}
\label{ch:stochastic_methods}
An alternative approach to approximate matrix inversion is to employ
Monte Carlo methods that sample a distribution with an expectation
value equivalent to that of the inverted operator. Such methods have
been in existence for decades with the earliest reference noted here a
manuscript published in 1950 by Forsythe and Leibler
\citep{forsythe_matrix_1950}. In their outline, Forsythe and Liebler
in fact credit the creation of this technique to J. Von Neumann and
S.M. Ulam some years earlier than its publication. In 1952 Wasow
provided a more formal explanation of Von Neumann and Ulam's method
\citep{wasow_note_1952} and Hammersley and Handscomb's 1964 monograph
\citep{hammersley_monte_1964} and Spanier and Gelbard's 1969 book
\citep{spanier_monte_1969} present additional detail on this topic
using a collection of references from the 1950's and early 1960's.

In this chapter, we will present the fundamentals of the Monte Carlo
method for discrete linear systems. Both the forward and adjoint
methods will be presented and analyzed including a brief variance
analysis of several Monte Carlo estimators. The Sequential Monte Carlo
and Monte Carlo Synthetic Acceleration methods will then be presented
as a means of leveraging the basic Monte Carlo methods in an iterative
refinement scheme that accelerates their convergence. Using these
iterative schemes, a set of modest variance reduction techniques will
be introduced and their effects on the Monte Carlo methods explored.

@@

%%---------------------------------------------------------------------------%%
\section{Direct Monte Carlo Method}
\label{sec:direct_mc}
In the context of matrix inversion, a direct (forward) method resembles an
adjoint Monte Carlo method in the reactor physics community where the
solution state is sampled and the source terms that contribute to it
are assembled. To achieve this, we build the direct method
Neumann-Ulam decomposition per Dimov's approach by first choosing a
probability matrix that is a row scaling of $\ve{H}$ such that its
components are:
\begin{equation}
  p_{ij} = \frac{|h_{ij}|}{\sum_j |h_{ij}|}\:.
  \label{eq:direct_probability}
\end{equation}
From this, we then see that the probability of transitioning from a
state $i$ to a state $j$ is implicitly linked to the original operator
$\ve{A}$ in that those terms with large values, and therefore those
that make the greatest contribution to the numerical solution, will be
sampled with a higher probability than smaller terms. In addition, the
row scaling provides a normalization over the state to which we are
transitioning such that $\sum_j p_{ij} = 1$, meaning that we sample
the probabilities over the rows of the matrix. The components of
the weight matrix are then defined by
Eq~(\ref{eq:neumann_ulam_decomposition}) as:
\begin{equation}
  w_{ij} = \frac{h_{ij}}{p_{ij}}\:.
  \label{eq:direct_weight}
\end{equation}
It should be noted here that if $\ve{A}$ is sparse, then $\ve{H}$,
$\ve{P}$, and $\ve{W}$ must be sparse as well by
definition. Additionally, we only compute $\ve{P}$ and $\ve{W}$ from
the non-zero elements of $\ve{H}$ as those components that are zero
will not participate in the random walk and thus produce identical
sparsity patterns for all matrices.

Using these matrices, we can then form the expectation value of the
direct solution. For a given random walk permutation $\nu$, we define
the weight of that permutation on the $m^{th}$ step to be:
\begin{equation}
  W_{m} = w_{i_0,i_1} w_{i_1,i_2} \cdots w_{i_{m-1},i_m}\:,
  \label{eq:direct_permutation_weight}
\end{equation}
such that the weight of each transition event contributes to the total
through multiplication with $W_0 = 1$ as the starting weight. The
contribution to the solution from a particular random walk permutation
with $k$ total events is then the \textit{forward estimator}:
\begin{equation}
  X_{i_0 = i}(\nu) = \sum_{m=0}^k W_{m} b_{i_m}\:,
  \label{eq:direct_permutation_contribution}
\end{equation}
where $X_{i_0 = i}(\nu)$ signifies that the solution state, $i_0$, in
which the random walk $\nu$ started is also the state, $i$, in which
we are tallying. We then define the probability that a particular
random walk permutation of $k$ events will occur:
\begin{equation}
  P_{\nu} = p_{i,i_1} p_{i_1,i_2} \cdots p_{i_{k-1},i_k}\:.
  \label{eq:direct_permutation_probability}
\end{equation}
Finally, we define the expectation value of $X$ to be the collection
of all random walk permutations and their probabilities:
\begin{equation}
  E\{X_i\} = \sum_{\nu} P_{\nu} X_{i}(\nu)\:,
  \label{eq:direct_expectation_value}
\end{equation}
which, if expanded, directly recovers the exact solution by forming
the Neumann series:
\begin{equation}
  \begin{split}
    E\{X_i\}
    &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
    w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{k-1},i_k} b_{i_k}\\ &= x_i\:,
  \end{split}
  \label{eq:direct_expectation_expansion}
\end{equation}
therefore providing an unbiased Monte Carlo estimator.

@@

\subsection{Direct Method: Evolution of a Solution}
\label{subsec:direct_evolution}
As a means of visually demonstrating the direct Monte Carlo method,
consider a 2-dimensional thermal diffusion problem with sources on the
left and right hand sides of the domain and a uniform source
throughout the domain of 1/5 the strength of the boundary sources as
shown in Figure~\ref{fig:heat_setup}.
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.2}{ \input{chapters/mc_background/heat_eq_setup.pdftex_t} }
  \end{center}
  \caption{\textbf{Problem setup for 2D heat equation.}
    \textit{Dirichlet conditions are set for the temperature on all 4
      boundaries of the Cartesian grid. Background source of 1/5 the
      value of the boundary sources present. $50 \times 50$ grid.}}
  \label{fig:heat_setup}
\end{figure}
For this problem, the number of histories used to compute the solution
at each grid point (state) in the domain was increased from 1 to 1000
in order to demonstrate the effects on the solution and the
statistical nature of the method. Figure~\ref{fig:direct_evolution}
gives these results.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/mc_background/direct_evolution.png}
  \end{center}
  \caption{\textbf{Direct Monte Carlo solution to the heat equation
      with varying numbers of histories.} \textit{Top left: 1 history
      per state. Top right: 10 histories per state. Bottom left: 100
      histories per state. Bottom right: 1000 histories per state.}}
  \label{fig:direct_evolution}
\end{figure}
As the number of histories used per state is increased, the
statistical variance of the solutions is decreased as more tallies are
made. Starting with single history at each state in the
domain, the high variance prevents a precise solution from being
obtained although we begin to see the solution take shape as
expected. At 1000 histories per state, enough tallies have been made
to generate a reasonable estimate for the structure of the
solution. It is interesting to note here that as the statistical
uncertainty is reduced at each grid point in the domain, the solution
is resolved in a certain sense, analagous to the convergence of a
traditional iterative method.
\clearpage

@@

Expanding the transition probabilities in the first sum again yields a
Neumman series in the same form as that explored for the forward
estimator, this time with the terms in reverse order and the
introduction of the Kronecker delta:
\begin{multline}
  \sigma_j^2 = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N} p_{i_k,i_{k-1}}\ldots p_{i_2,i_1} p_{i_1,i_0}
  w^2_{i_k,i_{k-1}}\ldots w^2_{i_2,i_1} w^2_{i_1,i_0}
  b^2_{i_0}\delta_{i_k,j} + \\ 2 \sum_{\nu} P_{\nu} \sum_{\substack{
      l=0 \\ m<l}}^k W_m W_l \delta_{i_m,j} \delta_{i_l,j} - x_j^2\:.
  \label{eq:collision_variance_2}
\end{multline}
If we again define $\ve{G} = \ve{P} \circ \ve{W} \circ \ve{W}$, we
then have:
\begin{multline}
  \sigma^2_j = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}g_{i_{k},i_{k-1}} \ldots
  g_{i_2,i_1}g_{i_1,i_0}b_{i_0}\delta_{i_k,j} + \\ 2 \sum_{\nu}
  P_{\nu} \sum_{\substack{ l=0 \\ m<l}}^k W_m W_l \delta_{i_m,j}
  \delta_{i_l,j} - x_j^2\:,
\label{eq:collision_variance_3}
\end{multline}
where now $\ve{G}$ in this case is the transpose of that used in
Eq~(\ref{eq:direct_variance_6}). The Kronecker delta again implies
that the variance contribution from each random walk will only be in
its current state and for many random walks, many starting weights of
$b_{i_0}$ will contribute to the variance in the same way as they
contribute to the solution tally. As the transpose is formed and the
Neumann series of $\ve{G}$ is in reverse order in
Eq~(\ref{eq:collision_variance_3}) relative to its formulation in the
forward estimator, then Eq~(\ref{eq:collision_variance_3}) and
Eq~(\ref{eq:direct_variance_6}) are equivalent and therefore the
collision estimator is bound by the same restrictions on
$\rho(\ve{G})$ and $\rho(\ve{H})$ to ensure that the variance is
finite.

@@

\subsection{Adjoint Method: Evolution of a Solution}
\label{subsec:adjoint_evolution}
As a means of visually demonstrating the adjoint Monte Carlo method,
again consider the 2-dimensional thermal diffusion problem with
sources on the left and right hand sides of the domain and a smaller
uniform source as shown in Figure~\ref{fig:heat_setup}. Using the
adjoint method with the collision estimator, the number of histories
sampled from the source was increased from 10 to 10,000,000 in order
to show its effects on the solution and the statistical nature of the
method. Figure~\ref{fig:adjoint_evolution} gives these results.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/mc_background/adjoint_evolution.png}
  \end{center}
  \caption{\textbf{Adjoint Monte Carlo solution to the heat equation
      with varying numbers of histories.} \textit{Top left: 10
      histories per state. Top right: 1,000 histories per
      state. Bottom left: 100,000 histories per state. Bottom right:
      10,000,000 histories per state.}}
  \label{fig:adjoint_evolution}
\end{figure}
As the number of histories used per state is increased, the
statistical variance of the solutions is decreased as more tally
contributions are made. At 10,000,000 histories per state, enough
information has been tallied to generate a reasonable estimate for the
structure of the solution. The visual difference between
Figures~\ref{fig:direct_evolution} and \ref{fig:adjoint_evolution} is
precisely that determined by their mathematics. As the adjoint
solution evolves with the addition of histories, more histories
emanate from the boundary and smaller uniform source with more
penetrating from the boundary into the domain and making contributions
to the tallies in those states.

%%---------------------------------------------------------------------------%%
\section{Sequential Monte Carlo}
\label{sec:sequential_mc}
The direct and adjoint Neumann-Ulam methods described are limited by a
convergence rate of $1/\sqrt{N}$ by the Central Limit Theorem where
$N$ is the number of random walk permutations. In 1962, Halton
presented a residual Monte Carlo method that moves towards exponential
convergence rates \citep{halton_sequential_1962} and further refined
his work some years later \citep{halton_sequential_1994}. Applications
of his work by the transport community have confirmed convergence
rates on the order of of $e^{-N}$ \citep{evans_residual_2003}. In
much the same way as projection methods, Halton's method, sequential
Monte Carlo, utilizes the adjoint Monte Carlo solver as a means of
directly reducing the elements residual vector. He proposed the
following iterative scheme as a solution to
Eq~(\ref{eq:linear_problem})\:
\begin{subequations}
  \begin{gather}
    \ve{r}^k = \ve{b} - \ve{A}\ve{x}^k\:,\\  
    \ve{A}\boldsymbol{\delta}^{k} = \ve{r}^{k}\:,\\
    \ve{x}^{k+1} = \ve{x}^k + \boldsymbol{\delta}^{k}\:,
  \end{gather}
  \label{eq:sequential_monte_carlo}
\end{subequations}
where the correction $\boldsymbol{\delta}^k$ is computed by the
adjoint Monte Carlo method at each iteration. The merits of Halton's
approach are immediately visible in that we have now broken the
binding of the convergence rate to the Central Limit Theorem. Here,
the Monte Carlo solver is used to produce a correction from the
residual, analogous to using the residual to extract a correction from
the search subspace in a projection method. By doing this, the Monte
Carlo error is bound in the correction used to update the solution and
therefore does not explicitly manifest itself in the overall
convergence of the solution. The downside of such a method is that if
the solution guess is poor, then many iterations are required in order
to reach exponential converge as the Monte Carlo error (and therefore
the Central Limit Theorem) does dominate in this situation.

%%---------------------------------------------------------------------------%%
\section{Monte Carlo Synthetic Acceleration}
\label{sec:mcsa}
Using the ideas of Halton, Evans and Mosher recently developed a Monte
Carlo solution method that was not prohibited severely by the quality
of the initial guess for the system \citep{evans_monte_2009} and later
applied it more rigorously as a solution mechanism for the radiation
diffusion equation \citep{evans_monte_2012}. With their new methods,
they achieved identical numerical results as conventional Krylov
solvers well as comparable performance in both number of iterations
and CPU time. Their approach was instead to use residual Monte Carlo
as a synthetic acceleration for a stationary method. To derive this
method, we begin by splitting the operator in
Eq~(\ref{eq:linear_problem})
\begin{equation}
  \ve{x} = (\ve{I} - \ve{A})\ve{x} + \ve{b}\:.
  \label{eq:linear_split}
\end{equation}
With this we can then define the stationary method
\textit{Richardson's iteration} as:
\begin{equation}
  \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,
  \label{eq:richardsons_iteration}
\end{equation}
which will converge if $\rho(\ve{I} - \ve{A}) < 1$. We then define the
solution error at the $k^{th}$ iterate relative to the true solution:
\begin{equation}
  \delta \ve{x}^k = \ve{x} - \ve{x}^k\:.
  \label{eq:mcsa_error}
\end{equation}
Subtracting Eq~(\ref{eq:richardsons_iteration}) from
Eq~(\ref{eq:linear_split}) we get:
\begin{equation}
  \delta \ve{x}^{k+1} = (\ve{I} - \ve{A})\delta \ve{x}^k\:.
  \label{eq:mcsa_setup_1}
\end{equation}
Subtracting from this $(\ve{I} - \ve{A})\delta \ve{x}^{k+1}$ yields:
\begin{equation}
  \begin{split}
    \ve{A}\delta \ve{x}^{k+1} &= (\ve{I} -
    \ve{A})(\ve{x}^{k+1}-\ve{x}^{k}) \\ &= \ve{r}^{k+1}\:.
    \label{eq:mcsa_setup_2}
  \end{split}
\end{equation}
Using this, we define the following scheme that will converge in one
iteration if $\ve{A}$ is inverted exactly:
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,\\
    \ve{A} \delta \ve{x}^{k+1} = \ve{r}^{k+1}\:,\\
    \ve{x} = \ve{x}^{k+1} + \delta \ve{x}^{k+1}\:.
  \end{gather}
  \label{eq:mcsa_setup_3}
\end{subequations}
However, $\ve{A}$ is only approximately inverted by our numerical
methods and therefore we instead pose an iterative scheme in which the
Monte Carlo solvers are used to invert the operator. The
\textit{Fixed-Point Monte Carlo Synthetic-Acceleration} (MCSA) method
is defined as:
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k\:,\\
    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\
    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,\\
    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
  \end{gather}
  \label{eq:mcsa}
\end{subequations}
where the adjoint Monte Carlo method is used to generate the solution
correction from the residual and Richardson's iteration in the first
step has been rewritten as a residual correction. Using Monte Carlo in
this way achieves the same effect as Halton's method, decoupling its
convergence rate from the overall convergence rate of the
method. Here, the approximate Monte Carlo solution is not driven to a
particular convergence as it merely supplies a correction for the
initial guess generated by Richardson's iteration. Rather, only a set
number of histories are required using the adjoint method to generate
the correction. In addition, the fact that the scheme in
Eq~(\ref{eq:mcsa_setup_3}) will converge in one iteration if $\ve{A}$
is inverted exactly means that as more and more stochastic histories
are used to compute the correction and the error is reduced towards
zero, the number of iterations required for MCSA to converge should
decrease accordingly, thus accelerating the solution.

>>>>

\chapter{Monte Carlo Synthetic Acceleration Methods}
\label{ch:stochastic_methods}
An alternative approach to approximate matrix inversion is to employ
Monte Carlo methods that sample a distribution with an expectation
value equivalent to that of the inverted operator. Such methods have
been in existence for decades with the earliest reference noted here a
manuscript published in 1950 by Forsythe and Leibler
\citep{forsythe_matrix_1950}. In their outline, Forsythe and Leibler
in fact credit the creation of this technique to J. Von Neumann and
S.M. Ulam some years earlier than its publication. In 1952 Wasow
provided a more formal explanation of Von Neumann and Ulam's method
\citep{wasow_note_1952} and Hammersley and Handscomb's 1964 monograph
\citep{hammersley_monte_1964} and Spanier and Gelbard's 1969 book
\citep{spanier_monte_1969} present additional detail on this topic
using a collection of references from the 1950's and early
1960's.

Following this work, Halton presented a residual Monte Carlo scheme
that improved dramatically on the performance of these methods
\citep{halton_sequential_1962}. Recent work in this area has utilized
these residual Monte Carlo methods in particle transport
\citep{evans_residual_2003}, giving accelerated convergence over
traditional methods. As an expansion of Halton's work, Evans and
others have leveraged these methods in a synthetic acceleration scheme
that has been shown to be competitive with production Krylov methods
for radiation transport problems and improve performance significantly
over Halton's sequential method \citep{evans_monte_2012}.

In this chapter, we will present the fundamentals of the Monte Carlo
method for discrete linear systems. Both the forward and adjoint
methods will be presented and analyzed including a brief variance
analysis of several Monte Carlo estimators. The Sequential Monte Carlo
method of Halton and Monte Carlo Synthetic Acceleration methods will
then be presented as a means of leveraging the basic Monte Carlo
methods devised by Neumann and Ulam in an iterative refinement scheme
that accelerates their convergence. Using these iterative schemes, a
set of modest variance reduction techniques will be introduced and
their effects along with other parameters of the algorithms on the
Monte Carlo solutions explored using a simple model transport problem.

@@

%%---------------------------------------------------------------------------%%
\section{Direct Monte Carlo Method}
\label{sec:direct_mc}
In the context of matrix inversion, a direct (forward) method
resembles an adjoint Monte Carlo method in the reactor physics
community where the solution state is sampled and the source terms
that contribute to it are assembled. To achieve this, we build the
forward Neumann-Ulam decomposition per Dimov's approach by first
choosing a probability matrix that is a row scaling of $\ve{H}$ such
that its components are:
\begin{equation}
  p_{ij} = \frac{|h_{ij}|}{\sum_j |h_{ij}|}\:.
  \label{eq:direct_probability}
\end{equation}
From this, we then see that the probability of transitioning from a
state $i$ to a state $j$ is implicitly linked to the original operator
$\ve{A}$ in that those terms with large values, and therefore those
that make the greatest contribution to the numerical solution, will be
sampled with a higher probability than smaller terms. In addition, the
row scaling provides a normalization over the state to which we are
transitioning such that $\sum_j p_{ij} = 1$, meaning that we sample
the probabilities over the rows of the matrix. The components of
the weight matrix are then defined by
Eq~(\ref{eq:neumann_ulam_decomposition}) as:
\begin{equation}
  w_{ij} = \frac{h_{ij}}{p_{ij}}\:.
  \label{eq:direct_weight}
\end{equation}
It should be noted here that if $\ve{A}$ is sparse, then $\ve{H}$,
$\ve{P}$, and $\ve{W}$ must be sparse as well by
definition. Additionally, we only compute $\ve{P}$ and $\ve{W}$ from
the non-zero elements of $\ve{H}$ as those components that are zero
will not participate in the random walk and thus produce identical
sparsity patterns for all matrices.

Using these matrices, we can then form the expectation value of the
forward solution. For a given random walk permutation $\nu$, we define
the weight of that permutation on the $m^{th}$ step to be:
\begin{equation}
  W_{m} = w_{i_0,i_1} w_{i_1,i_2} \cdots w_{i_{m-1},i_m}\:,
  \label{eq:direct_permutation_weight}
\end{equation}
such that the weight of each transition event contributes to the total
through multiplication with $W_0 = 1$ as the starting weight. The
contribution to the solution from a particular random walk permutation
with $k$ total events is then the \textit{forward estimator}:
\begin{equation}
  X_{i_0 = i}(\nu) = \sum_{m=0}^k W_{m} b_{i_m}\:,
  \label{eq:direct_permutation_contribution}
\end{equation}
where $X_{i_0 = i}(\nu)$ signifies that the solution state, $i_0$, in
which the random walk $\nu$ started is also the state, $i$, in which
we are tallying. We then define the probability that a particular
random walk permutation of $k$ events will occur:
\begin{equation}
  P_{\nu} = p_{i,i_1} p_{i_1,i_2} \cdots p_{i_{k-1},i_k}\:.
  \label{eq:direct_permutation_probability}
\end{equation}
Finally, we define the expectation value of $X$ to be the collection
of all random walk permutations and their probabilities:
\begin{equation}
  E\{X_i\} = \sum_{\nu} P_{\nu} X_{i}(\nu)\:,
  \label{eq:direct_expectation_value}
\end{equation}
which, if expanded, directly recovers the exact solution by forming
the Neumann series:
\begin{equation}
  \begin{split}
    E\{X_i\}
    &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
    w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{k-1},i_k} b_{i_k}\\ &= x_i\:,
  \end{split}
  \label{eq:direct_expectation_expansion}
\end{equation}
therefore providing an unbiased Monte Carlo estimator.

@@

\subsection{Direct Method: Evolution of a Solution}
\label{subsec:direct_evolution}
As a means of visually demonstrating the direct Monte Carlo method,
consider a 2-dimensional thermal diffusion problem with sources on the
left and right hand sides of the domain and a uniform source
throughout the domain of 1/5 the strength of the boundary sources as
shown in Figure~\ref{fig:heat_setup}.
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.2}{ \input{chapters/mc_background/heat_eq_setup.pdftex_t} }
  \end{center}
  \caption{\textbf{Problem setup for 2D heat equation.}
    \textit{Dirichlet conditions are set for the temperature on all 4
      boundaries of the Cartesian grid. Background source of 1/5 the
      value of the boundary sources present. $50 \times 50$ grid.}}
  \label{fig:heat_setup}
\end{figure}
For this problem, the number of histories used to compute the solution
at each grid point (state) in the domain was increased from 1 to 1000
in order to demonstrate the effects on the solution and the
statistical nature of the method. Figure~\ref{fig:direct_evolution}
gives these results.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/mc_background/direct_evolution.png}
  \end{center}
  \caption{\textbf{Direct Monte Carlo solution to the heat equation
      with varying numbers of histories.} \textit{Top left: 1 history
      per state. Top right: 10 histories per state. Bottom left: 100
      histories per state. Bottom right: 1000 histories per state.}}
  \label{fig:direct_evolution}
\end{figure}
As the number of histories used per state is increased, the
statistical variance of the solutions is decreased as more tallies are
made. Starting with a single history at each state in the domain, the
high variance prevents a precise solution from being obtained although
we begin to see the solution take shape as expected. At 1000 histories
per state, enough tallies have been made to generate a reasonable
estimate for the structure of the solution. It is interesting to note
here that as the statistical uncertainty is reduced at each grid point
in the domain, the solution is resolved in a certain sense, analogous
to the convergence of a traditional iterative method.
\clearpage

@@

Expanding the transition probabilities in the first sum again yields a
Neumann series in the same form as that explored for the forward
estimator, this time with the terms in reverse order and the
introduction of the Kronecker delta:
\begin{multline}
  \sigma_j^2 = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N} p_{i_k,i_{k-1}}\ldots p_{i_2,i_1} p_{i_1,i_0}
  w^2_{i_k,i_{k-1}}\ldots w^2_{i_2,i_1} w^2_{i_1,i_0}
  b^2_{i_0}\delta_{i_k,j} + \\ 2 \sum_{\nu} P_{\nu} \sum_{\substack{
      l=0 \\ m<l}}^k W_m W_l \delta_{i_m,j} \delta_{i_l,j} - x_j^2\:.
  \label{eq:collision_variance_2}
\end{multline}
If we again define $\ve{G} = \ve{P} \circ \ve{W} \circ \ve{W}$, we
then have:
\begin{multline}
  \sigma^2_j = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}g_{i_{k},i_{k-1}} \ldots
  g_{i_2,i_1}g_{i_1,i_0}b_{i_0}\delta_{i_k,j} + \\ 2 \sum_{\nu}
  P_{\nu} \sum_{\substack{ l=0 \\ m<l}}^k W_m W_l \delta_{i_m,j}
  \delta_{i_l,j} - x_j^2\:,
\label{eq:collision_variance_3}
\end{multline}
where now $\ve{G}$ in this case is the transpose of that used in
Eq~(\ref{eq:direct_variance_6}). The Kronecker delta again implies
that the variance contribution from each random walk will only be in
its current state and for many random walks, many starting weights of
$b_{i_0}$ will contribute to the variance in the same way as they
contribute to the solution tally. As the transpose is formed and the
Neumann series of $\ve{G}$ is in reverse order in
Eq~(\ref{eq:collision_variance_3}) relative to its formulation in the
forward estimator, then Eq~(\ref{eq:collision_variance_3}) and
Eq~(\ref{eq:direct_variance_6}) are equivalent and therefore the
collision estimator is bound by the same restrictions on
$\rho(\ve{G})$ and $\rho(\ve{H})$ to ensure that the variance is
finite.

@@

\subsection{Adjoint Method: Evolution of a Solution}
\label{subsec:adjoint_evolution}
As a means of visually demonstrating the adjoint Monte Carlo method,
again consider the 2-dimensional thermal diffusion problem with
sources on the left and right hand sides of the domain and a smaller
uniform source as shown in Figure~\ref{fig:heat_setup}. Using the
adjoint method with the collision estimator, the number of histories
sampled from the source was increased from 10 to 10,000,000 in order
to show its effects on the solution and the statistical nature of the
method. Figure~\ref{fig:adjoint_evolution} gives these results.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/mc_background/adjoint_evolution.png}
  \end{center}
  \caption{\textbf{Adjoint Monte Carlo solution to the heat equation
      with varying numbers of histories.} \textit{Top left: 10
      histories per state. Top right: 1,000 histories per
      state. Bottom left: 100,000 histories per state. Bottom right:
      10,000,000 histories per state.}}
  \label{fig:adjoint_evolution}
\end{figure}
As the number of histories used per state is increased, the
statistical variance of the solutions is decreased as more tally
contributions are made. At 10,000,000 histories per state, enough
information has been tallied to generate a reasonable estimate for the
structure of the solution. The visual difference between
Figures~\ref{fig:direct_evolution} and \ref{fig:adjoint_evolution} is
precisely that determined by their mathematics. As the adjoint
solution evolves with the addition of histories, more histories
emanate from the boundary the smaller uniform source in the domain
with more penetrating from the boundary into the domain and making
contributions to the tallies in those states as they are transported.

%%---------------------------------------------------------------------------%%
\section{Sequential Monte Carlo}
\label{sec:sequential_mc}
The direct and adjoint Neumann-Ulam methods described are limited by a
convergence rate of $1/\sqrt{N}$ by the Central Limit Theorem where
$N$ is the number of random walk permutations. In 1962, Halton
presented a residual Monte Carlo method that moves towards exponential
convergence rates \citep{halton_sequential_1962} and further refined
his work some years later \citep{halton_sequential_1994}. Applications
of his work by the transport community have confirmed convergence
rates on the order of of $e^{-N}$ \citep{evans_residual_2003}. In much
the same way as the projection methods outlined in
Appendix~\ref{ch:linear_problem}, Halton's method, sequential Monte
Carlo, utilizes the adjoint Monte Carlo solver as a means of directly
reducing the elements residual vector. He proposed the following
iterative scheme as a solution to Eq~(\ref{eq:linear_problem})\:
\begin{subequations}
  \begin{gather}
    \ve{r}^k = \ve{b} - \ve{A}\ve{x}^k\:,\\  
    \ve{A}\boldsymbol{\delta}^{k} = \ve{r}^{k}\:,\\
    \ve{x}^{k+1} = \ve{x}^k + \boldsymbol{\delta}^{k}\:,
  \end{gather}
  \label{eq:sequential_monte_carlo}
\end{subequations}
where the correction $\boldsymbol{\delta}^k$ is computed by the
adjoint Monte Carlo method at each iteration. The merits of Halton's
approach are immediately visible in that we have now broken the
binding of the convergence rate to the Central Limit Theorem. Here,
the Monte Carlo solver is used to produce a correction from the
residual, analogous to using the residual to extract a correction from
the search subspace in a projection method. By doing this, the Monte
Carlo error is bound in the correction used to update the solution and
therefore does not explicitly manifest itself in the overall
convergence of the solution. The downside of such a method is that if
the solution guess is poor, then many iterations are required in order
to reach exponential converge as the Monte Carlo error (and therefore
the Central Limit Theorem) does dominate in this situation.

%%---------------------------------------------------------------------------%%
\section{Monte Carlo Synthetic Acceleration}
\label{sec:mcsa}
Using the ideas of Halton, Evans and Mosher recently developed a Monte
Carlo solution method that was not prohibited severely by the quality
of the initial guess for the system \citep{evans_monte_2009} and later
applied it more rigorously as a solution mechanism for the radiation
diffusion equation \citep{evans_monte_2012}. With their new methods,
they achieved identical numerical results as conventional Krylov
solvers as well as comparable performance in both number of iterations
and CPU time. Their approach was instead to use residual Monte Carlo
as a synthetic acceleration for a stationary method. To derive this
method, we begin by splitting the operator in
Eq~(\ref{eq:linear_problem})
\begin{equation}
  \ve{x} = (\ve{I} - \ve{A})\ve{x} + \ve{b}\:.
  \label{eq:linear_split}
\end{equation}
With this we can then define the stationary method
\textit{Richardson's iteration} as:
\begin{equation}
  \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,
  \label{eq:richardsons_iteration}
\end{equation}
which will converge if $\rho(\ve{I} - \ve{A}) < 1$. We then define the
solution error at the $k^{th}$ iterate relative to the true solution:
\begin{equation}
  \delta \ve{x}^k = \ve{x} - \ve{x}^k\:.
  \label{eq:mcsa_error}
\end{equation}
Subtracting Eq~(\ref{eq:richardsons_iteration}) from
Eq~(\ref{eq:linear_split}) we get:
\begin{equation}
  \delta \ve{x}^{k+1} = (\ve{I} - \ve{A})\delta \ve{x}^k\:.
  \label{eq:mcsa_setup_1}
\end{equation}
Subtracting from this $(\ve{I} - \ve{A})\delta \ve{x}^{k+1}$ yields:
\begin{equation}
  \begin{split}
    \ve{A}\delta \ve{x}^{k+1} &= (\ve{I} -
    \ve{A})(\ve{x}^{k+1}-\ve{x}^{k}) \\ &= \ve{r}^{k+1}\:.
    \label{eq:mcsa_setup_2}
  \end{split}
\end{equation}
Using this, we define the following scheme that will converge in one
iteration if $\ve{A}$ is inverted exactly:
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,\\
    \ve{A} \delta \ve{x}^{k+1} = \ve{r}^{k+1}\:,\\
    \ve{x} = \ve{x}^{k+1} + \delta \ve{x}^{k+1}\:.
  \end{gather}
  \label{eq:mcsa_setup_3}
\end{subequations}
However, $\ve{A}$ is only approximately inverted by our numerical
methods and therefore we instead pose an iterative scheme in which the
Monte Carlo solvers are used to invert the operator. The
\textit{Fixed-Point Monte Carlo Synthetic-Acceleration} (MCSA) method
is defined as:
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k\:,\\
    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\
    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,\\
    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
  \end{gather}
  \label{eq:mcsa}
\end{subequations}
where a Neumann-Ulam Monte Carlo method is used to generate the
solution correction from the residual and Richardson's iteration in
the first step has been rewritten as a residual correction. Using
Monte Carlo in this way achieves the same effect as Halton's method,
decoupling its convergence rate from the overall convergence rate of
the method. Here, the approximate Monte Carlo solution is not driven
to a particular convergence as it merely supplies a correction for the
initial guess generated by Richardson's iteration. Rather, only a set
number of histories are required using the Neumann-Ulam method to
generate the correction. In addition, the fact that the scheme in
Eq~(\ref{eq:mcsa_setup_3}) will converge in one iteration if $\ve{A}$
is inverted exactly means that as more and more stochastic histories
are used to compute the correction and the error is reduced towards
zero, the number of iterations required for MCSA to converge should
decrease accordingly, thus accelerating the solution.
